---
title: "Getting Started with phenoQC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with phenoQC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(phenoQC)
```

## Introduction

phenoQC provides automated, reproducible quality control for phenotypic data
from plant breeding field trials. This vignette walks through a typical QC
session on the bundled `example_trial` dataset: a simulated 197-plot trial
with a handful of deliberate data quality issues baked in (duplicate plot
coordinates, missing values, an extreme yield outlier, an unreplicated
genotype, and a spatial gradient in plant height).

## Loading the example data

```{r load-data}
data(example_trial)
dim(example_trial)
#> [1] 197   8

head(example_trial)
#>   row col rep block genotype yield plant_height days_to_flower
#> 1   1   1   1     1   CHECK1  5.39         83.8             67
#> 2   1   2   1     1     G037  4.84         83.3             66
#> 3   1   3   1     1     G001  4.26         81.9             64
#> 4   1   4   1     1     G025  4.85         85.3             64
#> 5   1   5   1     1     G010  4.06         82.6             62
#> 6   1   5   1     1     G036  5.91         92.3             69
```

Notice rows 5 and 6 already show one of the injected issues: they share the
same `row`/`col` position (1, 5).

## Auto-detecting trait columns

Rather than typing out trait names by hand, `detect_trait_cols()` inspects
the numeric columns and excludes common structural fields like `row`, `col`,
`rep`, and `block`:

```{r detect-traits}
trait_cols <- detect_trait_cols(example_trial)
trait_cols
#> [1] "yield"          "plant_height"   "days_to_flower"
```

## Running the full QC pipeline

`phenoqc()` is the main entry point. It runs structure validation, missing
data analysis, outlier detection (IQR + spatial by default), and spatial
trend diagnostics in one call:

```{r run-phenoqc}
result <- phenoqc(example_trial, trait_cols = trait_cols)
#> ── phenoQC: Phenotypic Data Quality Control ──────────────────────────────
#>
#> ── Step 1: Validating trial structure ──
#>
#> ! 3 issue(s) found in trial structure.
#>
#> ── Step 2: Analysing missing data ──
#>
#> ℹ 5 missing value(s) across 3 trait(s).
#>
#> ── Step 3: Detecting outliers ──
#>
#> ℹ 5 outlier(s) flagged.
#>
#> ── Step 4: Checking spatial trends ──
#>
#> ! 1 trait(s) show spatial trend.
#>
#> ── QC Complete ──
#>
#> ✔ Use `summary()` for an overview or `qc_report()` to generate an HTML
#>   report.
```

The result is a `phenoqc_result` object with a `print()` method for a quick
overview:

```{r print-result}
result
#>
#> ── phenoQC Result
#> • Observations: 197
#> • Traits: yield, plant_height, days_to_flower
#> • Outliers: 6 total flagged
#> • Missing: 5 total
#> • Spatial: 1 trait(s) with trend
```

### Summary output

`summary()` returns a per-trait overview tibble:

```{r summary-result}
summary(result)
#> # A tibble: 3 × 7
#>   trait          n_obs n_missing n_outliers spatial_trend mean_raw sd_raw
#>   <chr>          <int>     <int>      <int> <lgl>            <dbl>  <dbl>
#> 1 yield            194         3          1 FALSE             5.02   1.26
#> 2 plant_height     195         2          1 TRUE             87.9    6.61
#> 3 days_to_flower   197         0          4 FALSE            64.7    2.42
```

`plant_height` is flagged with `spatial_trend = TRUE` and shows one outlier
even though its IQR-only outlier count is zero -- that's the combined
IQR + spatial flag at work, covered in more depth in the
`vignette("spatial-analysis")` article.

The full result also exposes each intermediate table individually:

```{r result-components}
result$validation
#> # A tibble: 4 × 4
#>   check                      status n_issues details
#>   <chr>                      <chr>     <int> <glue>
#> 1 Duplicate plot coordinates fail          2 2 duplicate position(s) detected
#> 2 Missing plot positions     warn          4 4 position(s) missing from grid
#> 3 Trait column types         pass          0 All trait columns are numeric
#> 4 Replication balance        warn          1 1 genotype(s) with unequal replication

result$missing
#> # A tibble: 3 × 4
#>   trait          n_total n_missing pct_missing
#>   <chr>            <int>     <int>       <dbl>
#> 1 yield              197         3         1.5
#> 2 plant_height       197         2         1
#> 3 days_to_flower     197         0         0
```

## Digging into individual checks

Each step of `phenoqc()` is also available as a standalone function, useful
for interactive exploration or building a custom pipeline.

### Duplicate plot coordinates

```{r check-duplicates}
qc_check_duplicates(example_trial)
#> # A tibble: 2 × 8
#>     row   col   rep block genotype yield plant_height days_to_flower
#>   <int> <int> <int> <int> <chr>    <dbl>        <dbl>          <int>
#> 1     1     5     1     1 G010      4.06         82.6             62
#> 2     1     5     1     1 G036      5.91         92.3             69
```

### Statistical (IQR) outliers

```{r outliers-statistical}
yield_out <- qc_outliers_statistical(example_trial, "yield", method = "iqr")
attr(yield_out, "outlier_summary")
#> # A tibble: 1 × 6
#>   trait method threshold n_total n_outliers pct_outliers
#>   <chr> <chr>      <dbl>   <int>      <int>        <dbl>
#> 1 yield iqr          1.5     194          1          0.5

yield_out[yield_out$yield_outlier, c("row", "col", "genotype", "yield", "yield_zscore")]
#>    row col genotype yield yield_zscore
#> 50   5  10     G009  18.5       10.731
```

That's the injected extreme yield value (18.5 t/ha against a trial mean
around 5 t/ha) -- easily caught by a simple IQR rule.

### Spatial outliers

```{r outliers-spatial}
yield_spatial <- qc_outliers_spatial(example_trial, "yield")
yield_spatial[
  yield_spatial$yield_spatial_outlier,
  c("row", "col", "genotype", "yield", "yield_spatial_residual")
]
#>    row col genotype yield yield_spatial_residual
#> 50   5  10     G009  18.5                    13.6
```

The same plot is flagged spatially too, with a large positive residual
relative to its field neighbours.

### Plotting outliers on the field layout

```{r plot-outliers, fig.alt = "Field heatmap of yield with the outlier plot marked with a red X"}
qc_plot_outliers(qc_flag_outliers(example_trial, "yield"), "yield")
```

`qc_plot_outliers()` draws a field heatmap of the raw trait values and
overlays an "X" marker on any plot flagged by `qc_flag_outliers()`.

## Generating an HTML report

Once you're happy with the QC run, `qc_report()` renders everything --
structure checks, missing data plots, outlier summaries, field heatmaps, and
spatial diagnostics -- into a single HTML report:

```{r qc-report}
qc_report(result, output_file = "example_trial_qc.html", open = FALSE)
#> ℹ Rendering QC report to 'example_trial_qc.html'...
#> ✔ Report saved to 'example_trial_qc.html'
```

Set `open = TRUE` (the default in interactive sessions) to have the report
open automatically in your browser.
