---
title: "Spatial Outlier Detection and Trend Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Outlier Detection and Trend Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(phenoQC)
data(example_trial)
```

## Why spatial methods?

Field trials are laid out in a grid, and neighbouring plots are rarely
independent: soil fertility gradients, moisture, shading, and management
effects all vary smoothly across a field. A purely statistical outlier
method like IQR or z-score treats every observation as exchangeable and
ignores this structure -- it can miss a plot that is perfectly "normal"
relative to the whole trial but starkly different from its immediate
neighbours (and it can also over-flag values that are extreme overall but
expected given a strong field trend).

phenoQC offers both:

- `qc_outliers_statistical()` -- IQR, z-score, or MAD outlier rules applied
  to the trait's overall distribution.
- `qc_outliers_spatial()` -- flags plots that deviate from the mean of
  their *k* nearest field neighbours.

This vignette shows a case in `example_trial` where the two methods
disagree, then covers the spatial trend and autocorrelation diagnostics.

## A case where spatial catches what IQR misses

`example_trial$plant_height` has a deliberate left-to-right spatial gradient
baked in (taller plants in higher-numbered columns), on top of the usual
plot-to-plot noise. Run both outlier methods on it:

```{r compare-methods}
ph_iqr <- qc_outliers_statistical(example_trial, "plant_height", method = "iqr")
ph_spatial <- qc_outliers_spatial(example_trial, "plant_height")

sum(ph_iqr$plant_height_outlier, na.rm = TRUE)
#> [1] 0

sum(ph_spatial$plant_height_spatial_outlier, na.rm = TRUE)
#> [1] 1
```

IQR finds nothing unusual -- every plant height value falls within a
plausible range for the trial as a whole. The spatial method flags one plot:

```{r spatial-only}
only_spatial <- which(ph_spatial$plant_height_spatial_outlier &
  !ph_iqr$plant_height_outlier)
example_trial[only_spatial, c("row", "col", "genotype", "plant_height")]
#>    row col genotype plant_height
#> 18   2   8     G031         79.3
```

At column 8, the field-wide gradient means most plots are quite tall
(column position pushes plant height up in this trial). A height of 79.3 cm
is unremarkable trial-wide, but it is well below what its immediate
neighbours in columns 7-9 are showing -- exactly the kind of local anomaly
`qc_outliers_spatial()` is designed to catch and that a global IQR rule,
by construction, cannot see.

Note the reverse also happens with `yield`: its single extreme value
(18.5 t/ha at row 5, col 10) is so far outside the trial's range that
*both* methods agree on it -- spatial methods complement statistical ones,
they don't replace them. In practice, `qc_flag_outliers()` (used internally
by `phenoqc()`) runs both and combines their flags.

## Detecting and visualising spatial trends

`qc_spatial_trend()` fits a 2D loess surface over the field coordinates and
reports how much of the trait's variance the spatial trend explains:

```{r spatial-trend}
trend_result <- qc_spatial_trend(example_trial, "plant_height")
attr(trend_result, "spatial_diagnostic")
#> # A tibble: 1 × 6
#>   trait        var_raw var_trend var_ratio has_trend interpretation
#>   <chr>          <dbl>     <dbl>     <dbl> <lgl>     <chr>
#> 1 plant_height    43.7      22.8     0.521 TRUE      Strong spatial trend detected
```

Over half of the raw variance in `plant_height` is explained by its
position in the field (`var_ratio = 0.521`) -- confirming the gradient we
know is there by design. Compare that to `yield`, which has no designed
spatial structure:

```{r spatial-trend-yield}
attr(qc_spatial_trend(example_trial, "yield"), "spatial_diagnostic")
#> # A tibble: 1 × 6
#>   trait var_raw var_trend var_ratio has_trend interpretation
#>   <chr>   <dbl>     <dbl>     <dbl> <lgl>     <chr>
#> 1 yield    1.58     0.103    0.0651 FALSE     No significant spatial trend
```

### Plotting raw values, trend, and residuals

`qc_plot_spatial()` facets the raw field layout, the fitted trend surface,
and the residuals side by side:

```{r plot-spatial, fig.alt = "Three-panel field heatmap of raw plant height, fitted spatial trend, and residuals"}
qc_plot_spatial(trend_result, "plant_height")
```

The raw and trend panels should look visibly similar (a left-to-right
colour gradient) if a genuine spatial trend is present; the residual panel
is where any remaining spatial anomalies -- like row 18 above -- become
easier to spot visually.

## Testing spatial autocorrelation with Moran's I

Beyond the loess-based trend test, `qc_spatial_autocorrelation()` computes
Moran's I, a formal statistic for whether nearby plots are more similar
than expected under spatial randomness.

```{r morans-i-plant-height}
qc_spatial_autocorrelation(example_trial, "plant_height")
#> $statistic
#> [1] 0.1311
#>
#> $expected
#> [1] -0.0052
#>
#> $p_value
#> [1] 0
#>
#> $interpretation
#> [1] "Significant positive spatial autocorrelation (clustering)"
```

A positive Moran's I well above the expected value under the null, with a
p-value of essentially 0, confirms significant spatial clustering in
`plant_height` -- consistent with the trend result above.

Compare with a trait that has no spatial structure:

```{r morans-i-yield}
qc_spatial_autocorrelation(example_trial, "yield")
#> $statistic
#> [1] -0.0091
#>
#> $expected
#> [1] -0.0052
#>
#> $p_value
#> [1] 0.7219
#>
#> $interpretation
#> [1] "No significant spatial autocorrelation"
```

```{r morans-i-days}
qc_spatial_autocorrelation(example_trial, "days_to_flower")
#> $statistic
#> [1] -0.0051
#>
#> $expected
#> [1] -0.0051
#>
#> $p_value
#> [1] 0.9998
#>
#> $interpretation
#> [1] "No significant spatial autocorrelation"
```

For both `yield` and `days_to_flower`, the observed Moran's I is close to
its expectation and the p-value is large -- no evidence of spatial
autocorrelation, matching the fact that neither trait was designed with a
field trend.

## Recommended workflow

1. Run `qc_spatial_autocorrelation()` per trait as a quick screening test.
2. For traits with significant autocorrelation, follow up with
   `qc_spatial_trend()` to quantify and visualise the trend, and consider
   whether it reflects real field variation that should be modelled (e.g.
   with a spatial mixed model) rather than treated as noise.
3. Always run `qc_outliers_spatial()` alongside a statistical method --
   `phenoqc()` does this by default via `outlier_methods = c("iqr",
   "spatial")` -- so that local anomalies aren't missed just because they
   look unremarkable at the whole-trial level.
